---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# exdex

[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/paulnorthrop/exdex?branch=master&svg=true)](https://ci.appveyor.com/project/paulnorthrop/exdex)
[![Coverage Status](https://codecov.io/github/paulnorthrop/exdex/coverage.svg?branch=master)](https://app.codecov.io/github/paulnorthrop/exdex?branch=master)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/exdex)](https://cran.r-project.org/package=exdex)
[![Downloads (monthly)](https://cranlogs.r-pkg.org/badges/exdex?color=brightgreen)](https://cran.r-project.org/package=exdex)
[![Downloads (total)](https://cranlogs.r-pkg.org/badges/grand-total/exdex?color=brightgreen)](https://cran.r-project.org/package=exdex)

## Estimation of the Extremal Index

### What does exdex do?

The extremal index $\theta$ is a measure of the degree of local dependence in the extremes of a stationary process.  The `exdex` package performs frequentist inference about $\theta$ using two types of methodology.  

One type ([Northrop, 2015](https://doi.org/10.1007/s10687-015-0221-5)) is based on a model that relates the distribution of block maxima to the marginal distribution of the data, leading to a semiparametric maxima estimator.  Two versions of this type of estimator are provided, following [Northrop, 2015](https://doi.org/10.1007/s10687-015-0221-5) and [Berghaus and Bücher, 2018](https://doi.org/10.1214/17-AOS1621).  A slightly modified version of the latter is also provided.  Estimates are produced using both disjoint and sliding block maxima, the latter providing greater precision of estimation. A graphical block size diagnostic is provided.  

The other type of methodology uses a model for the distribution of threshold inter-exceedance times ([Ferro and Segers, 2003](https://doi.org/10.1111/1467-9868.00401)).  Three versions of this type of approach are provided: the iterated weight least squares approach of [Süveges (2007)](https://doi.org/10.1007/s10687-007-0034-2), the $K$-gaps model of [Süveges and Davison (2010)](https://doi.org/10.1214/09-AOAS292) and a similar approach of [Holesovsky and Fusek (2020)](https://doi.org/10.1007/s10687-020-00374-3) that we refer to as D-gaps. For the $K$-gaps and $D$-gaps models the `exdex` package allows missing values in the data, can accommodate independent subsets of data, such as monthly or seasonal time series from different years, and can incorporate information from censored inter-exceedance times. Graphical diagnostics for the threshold level and the respective tuning parameters $K$ and $D$ are provided.

### A simple example

The following code estimates the extremal index using the semiparametric maxima estimators, for an example dataset containing a time series of sea surges measured at Newlyn, Cornwall, UK over the period 1971-1976.  The block size of 20 was chosen using a graphical diagnostic provided by `choose_b()`.

```{r}
library(exdex)
theta <- spm(newlyn, 20)
theta
summary(theta)
```

Now we estimate $\theta$ using the $K$-gaps model. The threshold $u$ and runs parameter $K$ were chosen using the graphical diagnostic provided by `choose_uk()`.

```{r}
u <- quantile(newlyn, probs = 0.60)
theta <- kgaps(newlyn, u, k = 2)
theta
summary(theta)
```

### Installation

To get the current released version from CRAN:

```{r installation, eval = FALSE}
install.packages("exdex")
```

### Vignette

See `vignette("exdex-vignette", package = "exdex")` for an overview of the package.

