---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# exdex

[![Travis-CI Build Status](https://travis-ci.org/paulnorthrop/exdex.svg?branch=master)](https://travis-ci.org/paulnorthrop/exdex) 
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/paulnorthrop/exdex?branch=master&svg=true)](https://ci.appveyor.com/project/paulnorthrop/exdex)
[![Coverage Status](https://codecov.io/github/paulnorthrop/exdex/coverage.svg?branch=master)](https://codecov.io/github/paulnorthrop/exdex?branch=master)

## Estimation of the Extremal Index

This package is under development: it is not ready for use.

### What does exdex do?

The extremal index \eqn{\theta} is a measure of the degree of local dependence in the extremes of a stationary process.  The `exdex` package performs frequentist inference about \eqn{\theta} using two types of methodology.  One type ([Northrop, 2015](https://doi.org/10.1007/s10687-015-0221-5)) is based on a model that relates the distribution of block maxima to the marginal distribution of the data.  The other type uses a model for the distribution of threshold inter-exceedance times ([Ferro and Segers, 2003](https://doi.org/10.1111/1467-9868.00401)).  Two versions of the the latter type of approach are provided, following [Süveges (2007)](https://doi.org/10.1007/s10687-007-0034-2) and
[Süveges and Davison (2010)](https://doi.org/10.1214/09-AOAS292).

### A simple example

### Installation

You can install exdex from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools") 
devtools::install_github("paulnorthrop/exdex")
```

### Vignette

See `vignette("exdex-vignette", package = "exdex")` for an overview of the package.

